---
title: "Pulses 2023 - drivers and barriers"
author: "Fie Langmann. ORCID: 0000-0002-7698-1146. Aarhus University, Department of Public Health"
date: "2024-12-05"
output: html_document
---

## This markdown includes code and output for factor analyses

<!-- The factor analysis relates to this article: *Article name here, doi:* -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE,message = FALSE)
options(width = 100)

#libraries
library(magrittr)
library(dplyr)
library(ltm)
library(psych)
library(tidyverse)
library(factoextra)
```

```{r data, include=FALSE}
data <- readr::read_csv("data/pulses23.csv")%>%
  dplyr::select(starts_with("q_17_"), starts_with("q_18_"), "q_1_1")
heavy <- data %>% filter(q_1_1 == "Heavy consumer")
light <- data %>% filter(q_1_1 == "Light consumer")

labels_Q17 <- readxl::read_excel("labels_Q17.xlsx",col_names = FALSE) %>% unlist() %>%  as.vector()
labels_Q18 <- readxl::read_excel("labels_Q18.xlsx",col_names = FALSE) %>% unlist() %>%  as.vector()

heavy_bar<- heavy %>% filter(!q_17_23==1) %>% dplyr::select(q_17_1:q_17_22)
heavy_dri<- heavy %>% filter(!q_18_26==1) %>% dplyr::select(q_18_1:q_18_25)
colnames(heavy_bar) <- labels_Q17
colnames(heavy_dri) <- labels_Q18

light_bar<- light %>% filter(!q_17_23==1) %>% dplyr::select(q_17_1:q_17_22)
light_dri<- light %>% filter(!q_18_26==1) %>% dplyr::select(q_18_1:q_18_25)
colnames(light_bar) <- labels_Q17
colnames(light_dri) <- labels_Q18
```

# Heavy users

## FA on tetrachoric correlations

-   Numbers of factor determined with screeplot. Factor should explain more than one variable (eigenvalue \> 1)
-   Factor analysis with minimum residual for extraction and Varimax for rotation.
-   Explained variance of total variance in

### Drivers

```{r heavy drivers FA, echo=FALSE}
# Simple scree plot
scree(tetrachoric(heavy_dri)$rho, pc = F) # 4 factors based on Eigenvalues < 1. Factor must explain more variance than one variable.

fa_result  <- tetrachoric(heavy_dri)$rho %>%
  fa(nfactors = 4,fm="wls",rotate = "varimax") #Varimax is orthogonal rotation generating factors that are uncorrelated
fa_result

# Extract loadings and convert to a data frame
loadings_melted <- fa_result$loadings %>%
  unclass() %>%
  as_tibble(rownames = "Variable") %>%
  rename_with(~ paste0(., " (", round(fa_result$Vaccounted[2, ] * 100, 1), "%)"), -Variable) %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading") %>% 
  filter(Loading >= 0.5) #Loadings cutoff

# Create the bar plot with loadings values inside the bars
ggplot(loadings_melted, aes(x = Variable, y = abs(Loading), fill = Loading)) +
  facet_wrap(~ Factor, nrow = 1) + # Place factors in separate facets
  geom_bar(stat = "identity") + # Create bars
  geom_text(aes(label = round(Loading, 2)), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3.5) + # Add loadings values inside bars
  coord_flip() + # Flip axes for horizontal labels
  scale_fill_gradient2(name = "Loadings",
                       high = "blue", mid = "white", low = "red",
                       midpoint = 0, guide = FALSE) +
  ylab("Loading") +
  theme_bw(base_size = 10) + # Black-and-white theme
  theme(strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) # Adjust facet label size

```

### Barriers

```{r heavy barriers FA, echo=FALSE}
# Simple scree plot
scree(tetrachoric(heavy_bar)$rho, pc = F) # 3 factors based on Eigenvalues < 1. Factor must explain more variance than one variable.

fa_result  <- tetrachoric(heavy_bar)$rho %>%
  fa(nfactors = 3,fm="wls",rotate = "varimax") #Varimax is orthogonal rotation generating factors that are uncorrelated
fa_result

# Extract loadings and convert to a data frame
loadings_melted <- fa_result$loadings %>%
  unclass() %>%
  as_tibble(rownames = "Variable") %>%
  rename_with(~ paste0(., " (", round(fa_result$Vaccounted[2, ] * 100, 1), "%)"), -Variable) %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading") %>% 
  filter(Loading >= 0.5) #Loadings cutoff

# Create the bar plot with loadings values inside the bars
ggplot(loadings_melted, aes(x = Variable, y = abs(Loading), fill = Loading)) +
  facet_wrap(~ Factor, nrow = 1) + # Place factors in separate facets
  geom_bar(stat = "identity") + # Create bars
  geom_text(aes(label = round(Loading, 2)), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3.5) + # Add loadings values inside bars
  coord_flip() + # Flip axes for horizontal labels
  scale_fill_gradient2(name = "Loadings",
                       high = "blue", mid = "white", low = "red",
                       midpoint = 0, guide = FALSE) +
  ylab("Loading") +
  theme_bw(base_size = 10) + # Black-and-white theme
  theme(strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) # Adjust facet label size
```

# Light users

## FA on tetrachoric correlations

-   Numbers of factor determined with screeplot. Factor should explain more than one variable (eigenvalue \> 1)
-   Factor analysis with minimum residual for extraction and Variamx for rotation.
-   Explained variance of total variance in

### Drivers

```{r light drivers FA, echo=FALSE}
# Simple scree plot
scree(tetrachoric(light_dri)$rho, pc = F) # 5 factors based on Eigenvalues < 1. Factor must explain more variance than one variable.

fa_result  <- tetrachoric(light_dri)$rho %>%
  fa(nfactors = 5,fm="wls",rotate = "varimax") #Varimax is orthogonal rotation generating factors that are uncorrelated
fa_result

# Extract loadings and convert to a data frame
loadings_melted <- fa_result$loadings %>%
  unclass() %>%
  as_tibble(rownames = "Variable") %>%
  rename_with(~ paste0(., " (", round(fa_result$Vaccounted[2, ] * 100, 1), "%)"), -Variable) %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading") %>% 
  filter(Loading >= 0.5) #Loadings cutoff

# Create the bar plot with loadings values inside the bars
ggplot(loadings_melted, aes(x = Variable, y = abs(Loading), fill = Loading)) +
  facet_wrap(~ Factor, nrow = 1) + # Place factors in separate facets
  geom_bar(stat = "identity") + # Create bars
  geom_text(aes(label = round(Loading, 2)), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3.5) + # Add loadings values inside bars
  coord_flip() + # Flip axes for horizontal labels
  scale_fill_gradient2(name = "Loadings",
                       high = "blue", mid = "white", low = "red",
                       midpoint = 0, guide = FALSE) +
  ylab("Loading") +
  theme_bw(base_size = 10) + # Black-and-white theme
  theme(strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) # Adjust facet label size

```

### Barriers

```{r light barriers FA, echo=FALSE}
# Simple scree plot
scree(tetrachoric(light_bar)$rho, pc = F) # 4 factors based on Eigenvalues < 1. Factor must explain more variance than one variable.

fa_result  <- tetrachoric(light_bar)$rho %>%
  fa(nfactors = 4,fm="wls",rotate = "varimax") #Varimax is orthogonal rotation generating factors that are uncorrelated
fa_result

# Extract loadings and convert to a data frame
loadings_melted <- fa_result$loadings %>%
  unclass() %>%
  as_tibble(rownames = "Variable") %>%
  rename_with(~ paste0(., " (", round(fa_result$Vaccounted[2, ] * 100, 1), "%)"), -Variable) %>%
  pivot_longer(-Variable, names_to = "Factor", values_to = "Loading") %>% 
  filter(Loading >= 0.5) #Loadings cutoff

# Create the bar plot with loadings values inside the bars
ggplot(loadings_melted, aes(x = Variable, y = abs(Loading), fill = Loading)) +
  facet_wrap(~ Factor, nrow = 1) + # Place factors in separate facets
  geom_bar(stat = "identity") + # Create bars
  geom_text(aes(label = round(Loading, 2)), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 3.5) + # Add loadings values inside bars
  coord_flip() + # Flip axes for horizontal labels
  scale_fill_gradient2(name = "Loadings",
                       high = "blue", mid = "white", low = "red",
                       midpoint = 0, guide = FALSE) +
  ylab("Loading") +
  theme_bw(base_size = 10) + # Black-and-white theme
  theme(strip.text = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) # Adjust facet label size
```
